Computational biology in the 21st century:\\The growth of compressive omics

### 2015-09-10 main.tex
Some changes, and reviewer comments included

\title{Computational biology in the 21st century:\\Algorithms that scale}
\numberofauthors{3}

@@

% YWY: Take spell-check lines out of figure axes labels! (TeraFLOPS, KiloBase)
\begin{figure*}[htb!]
\centering
\subfloat[Growth of genomic sequence data (orange) as compared with the combined power of the top-500 supercomputer list (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/moore.png}
\label{fig:moore}
}
\subfloat[Growth of genomic sequence data (orange) as compared with hard drive capacity (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/kryder.png}
\label{fig:kryder}
}
\caption[Sequencing vs. Storage]{(a) Moore's and (b) Kryder's laws contrasted with genomic sequence data}
\label{fig:growth}
\end{figure*}

@@

Computer scientists routinely exploit the structure of various data in
order to reduce time or space complexity.
In computational biology, this approach has implicitly served researchers well.
Now-classical approaches such as principal component analysis (PCA) reduce the 
dimensionality of data in order to simplify analysis and uncover salient 
features~\cite{berger2013computational}.
As another example, clever indexing techniques such as the Burroughs-Wheeler 
Transform (BWT) take advantage of aspects of sequence 
structure~\cite{berger2013computational} to speed up computation and save storage.
This Review focuses on cutting-edge algorithmic advances for dealing with the growth in 
biological data by explicitly taking advantage of its unique structure; algorithms for gaining novel biological insights are not its 
focus.

@@

Recent advances in metagenomic search tools have relied on two improvements over
BLASTX: indexing and alphabet reduction.
RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
collision-free hash table.
The alphabet reduction, as it is reversible, can be thought of as a form of
lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
alphabet, with offsets stored to recover the original sequence in the full
alphabet.
The hash-table provides an efficient index of the database to be searched.
DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
`shaped seeds' instead of simple $k$-mer seeds to index the database.
DIAMOND demonstrates search performance three to four orders of magnitude faster
than BLASTX, but it is still linear in the size of the database being searched.

@@

An area of great interest in integrating biological network data is in global multiple network 
alignment, which is based on network topology and conserved biological function (IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn}).
IsoRank's approach is similar to the PageRank algorithm~\cite{berger2013computational} 
but performs a random walk on the product graph of two networks;
IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
PageRank-Nibble algorithm~\cite{berger2013computational}.
NetworkBLAST~\cite{atias2012comparative} is another approach which searches for 
conserved paths and conserved interaction clusters.
Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
which defines a distance metric between two nodes based on random walks.
DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
functions when applied to standard approaches such as neighbor voting for function annotation.
Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
analogue of principle component analysis; DCA reduces the dimensionality of a
network in such a way that function annotation can be performed in the reduced 
space, significantly decreasing running time while still providing competitive
prediction accuracy.
Using a compressed representation of metabolic networks, 
Mongoose~\cite{chindelevitch2014exact} is able to 
identify and unblock many blocked reactions in current models without loss of 
accuracy.

@@

Beyond redundancy, however, another attribute of large biological data sets
stands out.
Any high-dimensional data set, such as amino acid sequences of a bounded 
length, can be described as inhabiting a vector space.
For example, given an alphabet of 20 amino acids and a length bounded at 
500\footnote{While the largest-known protein, titin, is over 33,000 amino acids 
in length, proteins are typically described as comprising \emph{domains} that 
evolve and fold independently; domains rarely exceed 500 amino acids.}, one 
could
enumerate $20^{500} \approx 3 \times 10^{650}$ distinct protein sequences, as 
of July 
2015, the NCBI's exhaustive `NR' database lists $6.9 \times 10^{7}$ sequences.
This surprisingly lower number is not simply due to undersampling; thanks to 
evolution, only those protein
sequences which exhibit useful biological function should survive, and most
random sequences of amino acids would not be expected to form stable structures.
Thus, not only do biological data exhibit redundancy, they also tend not to 
inhabit anywhere near the entire feasible space (Figure~\ref{fig:dataspace}).
It seems that physical laws have constrained the data to a particular subspace
of the enumerable Cartesian space.

@@

Cluster-based search, as exemplified by `compressive omics,' can perform
approximate search within a radius $r$ of a query $q$ on a database $D$ with 
fractal dimension $d$ and metric entropy $k$ at the scale $r_c$ in time
proportional to
\begin{displaymath}
    O\Bigg(
    \underbrace{k}_{\textrm{metric entropy}} +
    \overbrace{\left|B_D(q,r)\right|}^{\textrm{output size}}
    \underbrace{\left(\frac{r+2r_c}{r}\right)^d}_{\textrm{scaling factor}}
     \Bigg) .
\end{displaymath}

@@

Biological data sets exhibit redundancy, and are constrained to subspaces by
physical laws.
This combination results in low fractal dimension and low metric entropy 
relative to the size of the data set, which suggests that `compressive omics'
will provide the ability for computation to scale sublinearly with massively growing data.

@@

\begin{table}
\label{fractal}
\caption{Metric entropy ratio (ratio of clusters to entries in database) and
fractal dimension at typical search radii for four biological data sets.
NCBI's `NR' non-redundant protein sequence and `NT' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
\tabcolsep=0.11cm
\begin{tabular}{H{3.0cm}p{2.2cm}p{1.8cm}}
\hline
Data set & Metric entropy ratio & Fractal dimension \\        
\hline
Nucleotide~sequences (NCBI NT) & 7:1 & 1.5\\
\hline
Protein~sequences (NCBI NR) & 5:1 & 1.6\\
\hline
Protein~structure (PDB) & 10:1 & 2.5\\
\hline
Chemical~structure (PubChem) & 11:1 & 0.2\\
\hline
\end{tabular}
\end{table}

@@

The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST and caBLASTP, accelerates existing tools (in this case, read
mappers including BWA or Bowtie2) by allowing them to operate in a compressed
space, and relying on a coarse and a fine phase.
In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{berger2013computational}, conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as `seeds' for longer sequence matches.
CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each 
one to its neighbors within a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.

@@

@@

While this Review has focused on the importance of algorithms that leverage the
structure of data to cope with the increased throughput in computational 
biology, it is also essential that techniques continue to make better 
predictions and models that demonstrate improved fidelity to wet-lab 
experiments.

>>>>

\title{Computational biology in the 21st century:\\The growth of compressive 
omics}
\numberofauthors{3}

@@

\begin{figure*}[htb!]
\centering
\subfloat[Growth of genomic sequence data (orange) as compared with the combined power of the top-500 supercomputer list (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/moore.png}
\label{fig:moore}
}
\subfloat[Growth of genomic sequence data (orange) as compared with hard drive capacity (blue); y-axis is logarithmic]{
\includegraphics[width=2.8in]{assets/kryder.png}
\label{fig:kryder}
}
\caption[Sequencing vs. Storage]{(a) Moore's and (b) Kryder's laws contrasted with genomic sequence data}
\label{fig:growth}
\end{figure*}

@@

Computer scientists routinely exploit the structure of various data in
order to reduce time or space complexity.
In computational biology, this approach has implicitly served researchers well.
Now-classical approaches such as principal component analysis (PCA) reduce the 
dimensionality of data in order to simplify analysis and uncover salient 
features~\cite{berger2013computational}.
As another example, clever indexing techniques such as the Burrows-Wheeler 
Transform (BWT) take advantage of aspects of sequence 
structure~\cite{berger2013computational} to speed up computation and save storage.
This Review focuses on cutting-edge algorithmic advances for dealing with the growth in 
biological data by explicitly taking advantage of its unique structure; algorithms for gaining novel biological insights are not its 
focus.

\fbox{\parbox[t]{\linewidth}{

(16S, shaped seeds, blocked
reactions, chemogenomics, de Bruijn)
\textbf{Definitions}

\textbf{chemogenomics} This means this and that

\textbf{metagenomics} This means another thing and so on and so forth blah blah blah blah blah.

\textbf{metabolomics} And more.

\textbf{shotgun sequencing} And more.

\textbf{homology search} Determining the function, structure, or identity of a gene sequence by locating similar sequences within an annotated database.

\textbf{transcriptome} And more.

\textbf{FASTQ} A common plain-text format for biological sequence data.

\textbf{BLAST} And more.

\textbf{16S}

\textbf{de Bruijn graph}

}}

@@

Recent advances in metagenomic search tools have relied on two improvements over
BLASTX: indexing and alphabet reduction.
RapSearch2~\cite{zhao2012rapsearch2} relies on alphabet reduction and a 
collision-free hash table.
The alphabet reduction, as it is reversible, can be thought of as a form of
lossless compression; a 20-letter amino acid alphabet is mapped onto a smaller
alphabet, with offsets stored to recover the original sequence in the full
alphabet.
The hash-table provides an efficient index of the database to be searched.
DIAMOND~\cite{buchfink2014fast} also relies on alphabet reduction, but uses
`shaped seeds'---essentially, $k$-mers with wildcards at specific positions---instead of simple $k$-mer seeds to index the database.
DIAMOND demonstrates search performance three to four orders of magnitude faster
than BLASTX, but it is still linear in the size of the database being searched.

@@

An area of great interest in integrating biological network data is in global multiple network 
alignment, which is based on network topology and conserved biological function (IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn}).
IsoRank's approach is similar to the PageRank algorithm~\cite{berger2013computational} 
but performs a random walk on the product graph of two networks;
IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
PageRank-Nibble algorithm~\cite{berger2013computational}.
NetworkBLAST~\cite{atias2012comparative} is another approach which searches for 
conserved paths and conserved interaction clusters.
Diffusion state distance (DSD)~\cite{cao2013going} is a diffusion-based approach
which defines a distance metric between two nodes based on random walks.
DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
functions when applied to standard approaches such as neighbor voting for function annotation.
Diffusion component analysis (DCA)~\cite{cho2015diffusion} amounts to a network-based 
analogue of principle component analysis; DCA reduces the dimensionality of a
network in such a way that function annotation can be performed in the reduced 
space, significantly decreasing running time while still providing competitive
prediction accuracy.
Using a compressed representation of metabolic networks, 
Mongoose~\cite{chindelevitch2014exact} is able to 
identify and unblock many blocked reactions--those whose metabolites are out of 
balance---in current models without loss of accuracy.

@@

Beyond redundancy, however, another attribute of large biological data sets
stands out.
Far fewer biological sequences exist than could be enumerated, but even more so, those that exist tend to be highly similar to many others.
Thanks to 
evolution, only those genes
which exhibit useful biological function survive, and most
random sequences of amino acids would not be expected to form stable structures.
Since two human genomes differ on average by only 0.1\%, a collection of 1,000 human genomes contain less than twice the unique information of a single genome~\cite{loh2012compressive}.
Thus, not only do biological data exhibit redundancy, they also tend not to 
inhabit anywhere near the entire feasible space (Figure~\ref{fig:dataspace}).
It seems that physical laws---in this case, evolution---have constrained the data to a particular subspace of the enumerable Cartesian space.

@@

Cluster-based search, as exemplified by `compressive omics'---the use of compression to accelerate analysis---can perform
approximate search within a radius $r$ of a query $q$ on a database $D$ with 
fractal dimension $d$ and metric entropy $k$ at the scale $r_c$ in time
proportional to
\begin{displaymath}
    O\Bigg(
    \underbrace{k}_{\textrm{metric entropy}} +
    \overbrace{\left|B_D(q,r)\right|}^{\textrm{output size}}
    \underbrace{\left(\frac{r+2r_c}{r}\right)^d}_{\textrm{scaling factor}}
     \Bigg) ,
\end{displaymath}
where $B_D(q,r)$ refers to the set of points in $D$ contained within a ball of
radius $r$ about a point $q$.

@@

Biological data sets exhibit redundancy, and are constrained to subspaces by
physical laws; that is, the vast majority of enumerable sequences and 
structures do not exist because they are not advantageous (or at least, have 
not been selected for by evolution).
This combination results in low fractal dimension and low metric entropy 
relative to the size of the data set, which suggests that `compressive omics'
will provide the ability for computation to scale sublinearly with massively growing data.

@@

\begin{table}
\label{fractal}
\caption{Metric entropy ratio (ratio of clusters to entries in database) and
fractal dimension at typical search radii for four biological data sets. Metric entropy ratio gives an estimate of the acceleration of coarse search with respect to na\"ive search, and as long as fractal dimension is low, coarse search should dominate total search time.
NCBI's `NR' non-redundant protein sequence and `NT' non-redundant nucleotide sequence databases are from June, 2015. Protein Databank (PDB) is from July, 2015. PubChem is from October, 2013.}
\tabcolsep=0.11cm
\begin{tabular}{H{3.0cm}p{2.2cm}p{1.8cm}}
\hline
Data set & Metric entropy ratio & Fractal dimension \\        
\hline
Nucleotide~sequences (NCBI NT) & 7:1 & 1.5\\
\hline
Protein~sequences (NCBI NR) & 5:1 & 1.6\\
\hline
Protein~structure (PDB) & 10:1 & 2.5\\
\hline
Chemical~structure (PubChem) & 11:1 & 0.2\\
\hline
\end{tabular}
\end{table}

@@

The CORA read mapper~\cite{yorukoglu2015compressive} applies a mid-size $l$-mer based 
read-compression approach with a compressive indexing of the reference genome
(referred to as a homology table).
CORA, like caBLAST~\cite{loh2012compressive} and 
caBLASTP~\cite{daniels2013comprssive}, accelerates existing tools (in this 
case, read
mappers including BWA or Bowtie2) by allowing them to operate in a compressed
space, and relying on a coarse and a fine phase.
In contrast, short seed-clustering schemes, such as those used in Masai~\cite{siragusa2013fast} and 
MrsFAST~\cite{berger2013computational}, conceptually differ from CORA in that those 
schemes aim to accelerate only the seed-to-reference matching step.
Thus, there is a subsequent seed-extension step, which is substantially more 
costly and still needs to be performed for each read and mapping individually, 
even when seeds are clustered. 
Through its $l$-mer based read compression model, CORA is able to accelerate 
and 
achieve asymptotically sublinear scaling for both the seed-matching and 
seed-extension steps within coarse-mapping, which comprises the major bulk of 
the read-mapping computation.
Traditionally, $k$-mers refer to short substrings of fixed length (often, but 
not necessarily, a power of two) used as `seeds' for longer sequence matches.
CORA uses much longer $k$-mers (e.g., 33-64 nucleotides long), and links each 
one to its neighbors within a small Hamming or Levenshtein distance.
The term $l$-mer distinguishes these substrings from typically-short $k$-mers.

@@

The compressive omics approach can, in some cases, come at the cost of accuracy.
However, these cases are well-defined.
Compressive omics never results in false positives (with respect to the na\"ive 
search technique being accelerated), because the fine search phase applies the
same comparison to the candidates as the na\"ive approach.
Furthermore, when the distance function used for comparisons is a metric---more specifically, when it obeys the triangle inequality---false negatives will also
never occur.
However, in practice, non-metric distance functions are used, such as E-values
in BLAST or cosine distance in esFragBag, and thus false negatives can occur.
In practice, these error rates are low, and recall better than 90\% has been
demonstrated~\cite{loh2012compressive,daniels2013compressive,yu2015entropy}

@@

\textbf{rewrite paragraph}
While this Review has focused on the importance of algorithms that leverage the
structure of data to cope with the increased throughput in computational 
biology, it is also essential that techniques continue to make better 
predictions and models that demonstrate improved fidelity to wet-lab 
experiments.
